---
title: "ENVS_193DS_FINAL"
author: "Yaasiel Leos Ledesma"
date: "2026-03-17"
editor: visual
format:
  html:
    toc: true # use this to display table of contents
execute:
  warning: false # use this to make sure messages don't show up.
  message: false # use this to make sure warning don't show up
---

[GitHub repository](https://github.com/yaasiell/ENVS-193DS_winter-2026_final.git)

# Set Up

```{r}
#| label: Read in Packages and Data
#| message: false
#| warning: false
# Reading in Packages and Data
library(tidyverse) # data wrangling and plotting
library(here) # direct file paths to call from
library(janitor) # to examine and read in data
library(readxl) # reads excel files
library(DHARMa) # checks diagnostics for model using simulated residuals
# read in kelp data set
kelp <- read_csv(here("data","Annual_Kelp_All_Years_20250903.csv"))
# read in mountain plover data
mopl <- read_csv(here("data","MOPL_nest-site_selection_Duchardt_et_al._2019.csv"))

```

## Problem 1

### a. Transparent statistical methods
In part 1, my coworker used a linear regression statistical method because it compared one predictor variable as precipitation to predict a continuous response variable as flooded wetland area. In part 2 of the results, my coworker used a Kruskal-Wallis test because they compared a median flooded wetland area with more than 2 water year classification categories (5 total).

### b. Figure needed
I think that one figure that could accompany the statement for part 2 would be a box plot. The x-axis would show the predictor variable of 5  water year classification categories as wet, above normal, below normal, dry, and critical drought. The y-axis would show the response variable of flooded wetland area ($m^2$). Each individual box plot would show the median, interquartile range and the spread to compare distribution across classifications that align with the Kruskal-Wallis test. 

### c. Suggestions for rewriting

The flooded wetland area did not have evidence of having a relationship with precipitation. This would suggest that precipitation on its own is not a good predictor variable to predict the wetland flooding area in the San Joaquin River Delta (linear regression, test statistic = N/A, df = N/A, p = 0.11, $\alpha$ = significance level). 

The median flooded wetland area did not differ between the water year classifications. This would suggest that across the wet, above normal, below normal, dry, and critical drought years had similar median flooding conditions (Kruskal-Wallis test, test statistic = N/A, df = N/A, p = 0.12, $\alpha$ = significance level)

### d. Interpretation

Based on the Peterson et al. study and my coworkers analysis, one additional variable that could influence the flooded area of a wetland would be local wetland management action as a categorical predictor variable. This variable might influence wetland flooding because the paper states that wetland management could make decisions on  moving, retaining or restricting water on specific areas near the wetlands.

## Problem 2. Data visualization
### a. Cleaning and summarizing
```{r}
#| label: Cleaning and Summarizing
#| message: false
#| warning: false
# Clean and Summarize data frame
kelp_clean <- kelp |> # creating a new object
  clean_names() |> # changes column names to lowercase and underscores
  # filters to use only the 3 sites chosen
  filter(site %in% c("IVEE","NAPL","CARP")) |> 
  # creates a new column with the full site names chosen
  mutate(site_name = case_when(
  site == "NAPL" ~ "Naples", # data frame site column == site abbreviation ~ new named column
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
)) |>
  mutate(site_name = as_factor(site_name), # as_factor() to convert site name to a factor 
         # fct_relevel() reorders the levels of a factor 
       site_name = fct_relevel(site_name,
                               "Naples", "Isla Vista", "Carpinteria")) |>
  # filter() used to filter and remove a specific observation
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |>
  group_by(site_name, year) |> # group_by() groups data by site and year for summarizing
  summarize(mean_fronds = mean(fronds)) |> # calculates the mean number of fronds for each site for each year
  ungroup() # removes the grouping after being summarized
slice_sample(kelp_clean, n = 5) # displays 5 rows from kelp_clean
str(kelp_clean) # str() displays the structure of the data set

```

### b. Understanding the data
The observation from 2000-09-22 in transect 6, quad 40 were filtered out because it had a frond value of -99999 which was used to represent missing data instead of an actual measurement.In the Methods and Protocls section of "KFCD abundance and size of giant kelp" data set in the SBC LTER's data catalog it indicates that the data was either not collected for a particular species or the datasheet was lost needing to be removed before being analyzed. 

### c. Visualize the data
```{r}
#| label: Visualizing Kelp data
#| message: false
#| warning: false
# Visualizing Kelp Data
# ggplot() to initialize objeect
ggplot(kelp_clean, aes(x = year, # calls for year to represent x-axis
                       y = mean_fronds, # calls for mean fronds data to represent y-axis
                       color = site_name)) + # color added to site name
  # geom_line() adds a line to connect mean frond values for each site
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) + # geom_point() adds points for each yearly mean frond value
  labs( # labs() calls for labels 
    title = "Sites vary in mean kelp fronds per year", # labels the title
    x = "Year", # labels the x-axis
    y = "Mean kelp fronds per year", # labels the y-axis
    color = "Site" # labels the legend 
  ) +
  # adds the required title and axis/legend text
  scale_color_manual( # manually assing colors to values
    values = c(
      "Naples" = "dodgerblue2", # change the color of Naples data
      "Isla Vista" = "darkgreen", # change the color of Isla Vista data
      "Carpinteria" = "lightpink3" # change the color of Carpinteria data
    )
  ) +
  # applies a minimal theme with a custom font that is consistent
  theme_minimal(base_family = "Arial") +
  theme(
    plot.title = element_text(hjust = 0.5), # adjust the centering of the plot title
    panel.border = element_blank() # removes the panel boarder to show a grid without a border 
  )
```

### d. Interpretation
The site with the most variable mean kelp fronds per year was found to be Isla Vista because of how the graph shows large fluctuating peaks near 30-40 and lower peaks at 5-10 fronds representing a wide range of variability across multiple years. The site with the least variable mean kelp frond per year was found to be Carpinteria because the values stay at a consistent range throughout the years with a smaller fluctuation from 4-15 fronds per year. 

## Problem 3. Data analysis

### a. Response variable 
The 0 and 1's in the data set represent the response variable if a potential nest site was used by a Mountain Plover. A value of 1 represents that a nest site was "used" by the Mountain Plover shown in the id column, and a value of 0 means that the site was "available" or unused by a Mountain Plover shown in the id column.

### b. Purpose of study 
For the first variable I found to be Visual obstruction as a continuous variable because it measured vegetation height and density around the nesting site using a Robel pole. As the visual obstruction increases, the probability of a Mountain Plover nest site being used would decrease because they prefer live in areas with lower vegetation so they can see predators better. This information was found in the Methods Adults Mountain Plover Density in the Data collection section were visual obstruction is described as being a measure of vegitation structure. 

For the second predictor variable of Distance to prairie dog colony edge it is also a continuous variable that is meant to represent how far a nesting location is from from a prairie colony. As the distance from the colony edge increases, the probability that a Mountain Plover nesting site would decrease or vary depending on habitat conditions  because the prairie dog would create an open habitat that the Plover might like. This information was found in the Methods Modeling framework section in table 1 that shows the distance to colony edge in the Prairie dog colony-direct section. 

### c. Table of models
| Model | Visual Obstruction (vo_nest) | Distance to Colony Edge (edgedist) | Model Description |
|------|------------------------------|------------------------------------|------------------|
| 1 | No | No | Null model (intercept only) |
| 2 | Yes | No | Model with visual obstruction only |
| 3 | No | Yes | Model with distance to colony edge only |
| 4 | Yes | Yes | Full model with both predictors |

### d. Run the models
```{r}
#| label: Running Models
#| message: false
#| warning: false
# Running the models
mopl_clean <- mopl |> # created a new object 
  clean_names() |> # changes column names to lowercase and underscores
  # mutate to change the values to bianary numbers were 1 = used nest site, 0 = available
  mutate(used_bin = if_else(id == "used", 1, 0)) |>
  select(used_bin, edgedist, vo_nest) # select() keeps only the columns selected to be used as response and predictor variables

```

```{r}
#| label: Running the models
#| message: false
#| warning: false
#Running the Models

# null model: includes only the intercept (no predictors)
m1 <- glm(used_bin ~ 1, data = mopl_clean, family = "binomial")
# model with visual obstruction as the only predictor
m2 <- glm(used_bin ~ vo_nest, data = mopl_clean, family = "binomial")
# model with distance to prairie dog colony edge as the only predictor
m3 <- glm(used_bin ~ edgedist, data = mopl_clean, family = "binomial")
# full model: includes both predictors
m4 <- glm(used_bin ~ vo_nest + edgedist, data = mopl_clean, family = "binomial")

```

### e. Select the best model

The best model as determined by Akaike's Information Criterion (AIC) was the visual obstruction at the nest as the only predictor of the probability of Mountain plover nest site usage because it had the lowest AIC value compared to all other models.

### f. Check the diagnostics
```{r}
#| label: Checking Diagnostics for model
# checking diagnostics for the model I selected

# simulates residuals based on fitted model (m2)
sim_res <- simulateResiduals(m2)
# visualize diagnostic plots that includes QQ test, dispersion test, and outlier test
plot(sim_res)
```


### g. Visualize the model predictions
### h. Write a caption for your figure
### i. Calculate model predictions
### j. Interpret your results

## Problem 4. Affective and exploratory visualizations
### a. Comparing visualizations
### b. Designing an analysis
### c. Check any assumptions and run your analysis
### d. Create a visualization
### e. Write a caption
### f. Write about your results 
### g. Sharing your affective visualization


